---
title: "Predicting 2024 AL & NL Home Run Leaders"
author: "Jacob Hallowell"
date: "2024-03-07"
output: html_notebook
---

I am conducting this experiment for a couple reasons. To be honest, I saw FanDuel had a section where I can bet on who will lead the American and National Leagues in Home Runs for the upcoming season. I figured I can use publicly available data to get a good idea for who it will be, rather than using my gut.

Let's start by loading in the datasets and separating them so I can join them later.
```{r}
# clear the environment
rm(list = ls())

library(tidyverse)
library(caret)
library(dplyr)
library(data.table)
library(Hmisc)
library(randomForest)

batter2023 <- read.csv("batter2023.csv")
batter2022 <- read.csv("batter2022.csv")
batter2021 <- read.csv("batter2021.csv")
```

After doing this I want to get subsets of the data so I can join them to other datasets. This way I can predict how many home runs they will hit. Unfortunately, I only grabbed stuff from baseball savant. There are going to be flaws with this, but I am going to point them out as I go along.
```{r}
home_runs2023 <- batter2023 %>%
  select(player_id, home_run)
home_runs2022 <- batter2022 %>%
  select(player_id, home_run)
home_runs2021 <- batter2021 %>%
  select(player_id, home_run)

IDtoName21 <- batter2021 %>%
  select(ï..last_name..first_name, player_id)
IDtoName22 <- batter2022 %>%
  select(ï..last_name..first_name, player_id)
IDtoName23 <- batter2023 %>%
  select(ï..last_name..first_name, player_id)
```

After this I really want to clean the data I have now. In my mind I want the columns to be similar so the machine has less trouble interpreting the columns. With how I downloaded the data, the percentage columns are out of 100 and the other columns are not. For example, ground ball percent is 24.1 not 0.241 like the rest.
```{r}
convert_to_decimal <- function(df, cols) {
  for (col in cols) {
    df[[col]] <- df[[col]] / 100
  }
  return(df)
}

percentage_columns <- c('groundballs_percent', 'flyballs_percent', 'linedrives_percent', 'k_percent', 'bb_percent', 'sweet_spot_percent', 'barrel_batted_rate', 'hard_hit_percent', 'whiff_percent', 'swing_percent', 'pull_percent', 'straightaway_percent', 'opposite_percent')

# Convert for each dataset
batter2021 <- convert_to_decimal(batter2021, percentage_columns)
batter2022 <- convert_to_decimal(batter2022, percentage_columns)
batter2023 <- convert_to_decimal(batter2023, percentage_columns)
```

The next step is joining the home runs for the next year into the data. I can do this by joining. I also need to rename the future home runs.
```{r}
home_runs2022 <- home_runs2022 %>% rename(next_year_home_run = home_run)
home_runs2023 <- home_runs2023 %>% rename(next_year_home_run = home_run)

data2021 <- left_join(batter2021, home_runs2022, by = "player_id")
data2022 <- left_join(batter2022, home_runs2023, by = "player_id")
```

Now we can start cleaning our data for model building.
```{r}
# remove NAs in data 2021-2022
data2021 <- data2021 %>% filter(next_year_home_run >= 0)
data2022 <- data2022 %>% filter(next_year_home_run >= 0)

columns_to_remove <- c('player_id', 'year', 'ï..last_name..first_name')

# Remove specified columns from each dataset
data2021 <- data2021[, !(names(data2021) %in% columns_to_remove)]
data2022 <- data2022[, !(names(data2022) %in% columns_to_remove)]
data2023 <- batter2023[, !(names(batter2023) %in% columns_to_remove)]
```

Now that I have my cleaned data I can do a couple of things. First I want to create a correlation matrix and then I want to make a model. Lets start with making a model then I will show I correlation matrix. This is going to be a relatively simple model because I don't want to get too complex yet.
```{r}
combined_data <- rbind(data2021, data2022)

library(caTools)
set.seed(42) # for reproducibility
split <- sample.split(combined_data$next_year_home_run, SplitRatio = 0.8)
training_set <- subset(combined_data, split == TRUE)
testing_set <- subset(combined_data, split == FALSE)

# i love random forest
library(randomForest)
set.seed(42)
model <- randomForest(next_year_home_run ~ ., data = training_set, ntree = 100)

predictions <- predict(model, testing_set)
mse <- mean((testing_set$next_year_home_run - predictions)^2)
print(mse)


library(Metrics)
library(performance)
mae <- mae(testing_set$next_year_home_run, predictions)

print(mae)

```

Now it is time to make predictions on our 2023 dataset.
```{r}
predictions2024 <- predict(model, data2023)
```

I want to attach my ids to my predictions this way I can actually see who is predicted for the most.
```{r}
#predictions2024 <- as.data.frame(predictions2024)

batter2023$PredHR <- predictions2024
final_pred <- batter2023 %>%
  select(ï..last_name..first_name, PredHR)
```

Lets take a look:
```{r}
final_pred %>%
  arrange(desc(PredHR)) %>%
  head()
```

Now I want to check out the correlation matrix. I want to use combined_data for that because it will be using actual data, not forecasted data.
```{r}
cor_matrix <- cor(combined_data[, sapply(combined_data, is.numeric)])
cor_with_next_year_homeruns <- cor_matrix["next_year_home_run", ]
sorted_correlations <- sort(cor_with_next_year_homeruns, decreasing = TRUE)
print(sorted_correlations)
```

Now I want to create a colored table
```{r}
library(DT)
cor_df <- as.data.frame(cor_matrix)

# Add row names as a new column for display
cor_df$Variable <- row.names(cor_df)
cor_table <- cor_df %>%
  select(Variable, next_year_home_run)

# Create the datatable and apply coloring
datatable(cor_df, options = list(pageLength = 5, autoWidth = TRUE)) %>%
  formatStyle(
    'Variable',
    target = 'row', # Highlight entire row
    backgroundColor = styleInterval(c(-0.5, 0.5), c('red', 'white', 'blue'))
  ) %>%
  formatStyle(
    names(cor_df),
    backgroundColor = styleInterval(c(-0.5, 0.5), c('red', 'white', 'blue')),
    color = 'black', # Text color
    fontWeight = 'bold'
  )

```

I dont like the whole table, so i am going to try and accomplish this with one row.
```{r}
cor_next_year_home_run <- cor_matrix["next_year_home_run", ]
cor_df_next_year_home_run <- as.data.frame(cor_next_year_home_run)
cor_df_next_year_home_run$Variable <- rownames(cor_df_next_year_home_run)

library(DT)

datatable(cor_df_next_year_home_run, options = list(pageLength = 5, autoWidth = TRUE)) %>%
  formatStyle(
    names(cor_df_next_year_home_run),
    backgroundColor = styleInterval(c(-0.5, 0.5), c('blue', 'white', 'red')),
    color = 'black', # Text color
    fontWeight = 'bold'
  )
```

